---
title: 'Exercise Sheet 10: Hypothesis Testing'
author: ''
date: ''
output:
  pdf_document:
    toc: true
    toc_depth: '3'
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
subtitle: "One-sample procedures (z / t / chi-square, proportions) + selected two-sample
  and chi-square methods"
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
set.seed(42)
```

# Quick reference: hypothesis testing building blocks

## Core objects

-   **Null hypothesis** $H_0$: baseline claim about a parameter (e.g.
    $\mu=\mu_0$).
-   **Alternative** $H_1$: departures from $H_0$ we want to detect
    (two-sided or one-sided).
-   **Test statistic** $W$: a function of the sample data.
-   **Rejection region** $R$: reject $H_0$ iff $W_{\text{obs}}\in R$.
-   **Type I error**: reject $H_0$ although $H_0$ is true. Probability
    $\alpha$.
-   **Type II error**: fail to reject $H_0$ although $H_1$ is true.
    Probability $\beta$.
-   **Power**: $1-\beta$.

## Quantiles and p-values in R

-   Normal critical value: `qnorm(1 - alpha/2)` and CDF: `pnorm()`
-   t critical value: `qt(1 - alpha/2, df)` and CDF: `pt()`
-   chi-square quantile: `qchisq(p, df)` and CDF: `pchisq()`

## Common tests in R (and what they correspond to)

-   One-sample mean (unknown $\sigma$): `t.test(x, mu = mu0)`
-   One-sample mean (known $\sigma$): compute $Z$ and use `pnorm()`
-   One-sample variance (normal model): compute $Q$ and use `pchisq()`
-   One-sample proportion (approx):
    `prop.test(x, n, p = p0, correct = FALSE)`
-   One-sample proportion (exact): `binom.test(x, n, p = p0)`
-   Two-sample means (Welch): `t.test(x, y)` (default is Welch)
-   Two-sample means (pooled, equal variances):
    `t.test(x, y, var.equal = TRUE)`
-   F test for equal variances (normal): `var.test(x, y)`
-   Chi-square test of independence (contingency table):
    `chisq.test(tab)`
-   Wilcoxon–Mann–Whitney (rank-based): `wilcox.test(x, y)`
-   Paired t-test: `t.test(x, y, paired = TRUE)`

------------------------------------------------------------------------

# Part A — Logic of hypothesis testing (conceptual)

## Exercise 1 (conceptual): $H_0$, $H_1$, statistic, rejection region

We use `mtcars$mpg` as a sample.

```{r}
data(mtcars)
x <- mtcars$mpg
n <- length(x)
c(n = n, mean = mean(x), sd = sd(x))
```

**Tasks**

1.  Write down $H_0$ and $H_1$ for each case (in symbols and one short
    sentence):

    -   

        (a) Two-sided: “the mean mpg equals 20”.
        H0: The mean mpg equals 20
        H1: The mean mpg does not equal 20

    -   

        (b) Right-tailed: “the mean mpg is larger than 20”.
        H0: The mean mpg is smaller or equal to 20 
        H1: The mean mpg is larger than 20

    -   

        (c) Left-tailed: “the mean mpg is smaller than 20”.
        H0: The mean mpg is larger or equal to 20 
        H1: The mean mpg is smaller than 20


2.  For the one-sample $t$-test, the test statistic is 
    $$
    T=\frac{\bar X-\mu_0}{S/\sqrt{n}}.
    $$
    Explain (in 2–3 sentences) what “large” values of $T$ mean in words.

    A: A large value for T indicates that he sample mean X is far away from the
    actual mean, arguing against H0.

3.  What is a **rejection region** $R$ for a two-sided test at level
    $\alpha=0.05$? (Answer using a critical value $t_{n-1,0.975}$.)
    
    $$R = \{T: T < -t_{n-1, 0.975}\text{ or }T > t_{n-1, 0.975}\}$$

4.  In 1–2 sentences each: define Type I error, Type II error, and
    power.
    
    Type I Error: Falsely rejecting $H_0$
    
    Type II Error: Failing to reject $H_0$ when it should be.
    
    Power: Probability of correctly rejecting $H_0$

------------------------------------------------------------------------

# Part B — One-sample tests in the normal model

## Exercise 2: One-sample $t$-test for a mean (unknown $\sigma$) + p-value

We test the mean mpg in `mtcars`.

Let $\mu_0=20$ and $\alpha=0.05$.

**Tasks**

1.  Compute $\bar x$, $s$, $n$, and the observed test statistic
    $$
    t_{\text{obs}}=\frac{\bar x-\mu_0}{s/\sqrt{n}}.
    $$

```{r}
meanx <- mean(mtcars$mpg)
s <- sd(mtcars$mpg)
n <- length(mtcars$mpg)
```

2.  Compute the **two-sided p-value** manually using the
    $t$-distribution:
    $$
      p = 2\,P\left(T\ge |t_{\text{obs}}|\right),\quad T\sim t_{n-1}.
    $$

```{r}
t_obs <- (meanx - 20)/(s / sqrt(n))
p_2sided <- 2 * (1 - pt(abs(t_obs), df = n-1))
```

3.  Check with `t.test()` and confirm that the p-value matches.

```{r}
t.test(mtcars$mpg)
```

4.  State your decision at $\alpha=0.05$: reject or not reject $H_0$?
    (Write 1–2 sentences: include p-value and conclusion in context.)

    The p-value is ~0.9328, which is much larger then 0.05. $\rightarrow$ $H_0$ cannot be rejected. $\rightarrow$ There is no signifitant evidence that the mean of mpg differs from 20.

------------------------------------------------------------------------

## Exercise 3: CI–test duality for the mean

Use the same data and $H_0:\mu=\mu_0$ with $\mu_0=20$ and $\alpha=0.05$.

**Tasks**

1.  Compute the **95% confidence interval** for $\mu$ using `t.test()`.

```{r}
ttest <-t.test(mtcars$mpg)
ttest$conf.int[1:2]
```

2.  Verify CI–test duality: check whether $\mu_0$ lies inside the CI.
    Explain why this agrees with your test decision in Exercise 2.

    $\mu_0$ lies inside the CI. This is because both the calculation for the CI aswell 
    as the calculation for $\mu_0$ use the same logic / formula, just isolated on another value.
    Because you can also derive from $\mu_0 \in$ CI that we cannot drop $H_0$, the same result
    as the t-test. This shows the duality of both approaches.
    
    
3.  Repeat (conceptually only): what happens to the decision rule if you
    change the confidence level from 95% to 99%?

    The CI grows.

------------------------------------------------------------------------

## Exercise 4: One-sample z-test for a mean (known $\sigma$) (illustration)

In practice $\sigma$ is rarely known. Here we do an **illustration**
using `airquality$Ozone`.

Assume (for this exercise) that the population standard deviation is
known and equals $\sigma_0=20$. We test $H_0:\mu=40$ vs $H_1:\mu\neq 40$
at $\alpha=0.05$.

```{r}
x <- airquality$Ozone
sigma <- 20
mu_0 <- 40
alpha <- .05
```

**Tasks**

1.  Compute the z-statistic
    $$
    z_{\text{obs}}=\frac{\bar x-\mu_0}{\sigma_0/\sqrt{n}}
    $$ with $\mu_0=40$ and $\sigma_0=20$.

```{r}
z_obs <- (mean(x, na.rm=T) - mu_0) / (sigma / sqrt(length(x)))
z_obs
```

2.  Compute the two-sided p-value using the standard normal
    distribution.

```{r}
p_2sided <- 2 * (1 - pnorm(abs(z_obs)))
p_2sided
```

3.  Compute the critical value $z_{0.975}$ and decide whether to reject.

```{r}
z_crit <- qnorm(1 - alpha / 2)
z_crit
#-> I do not reject H_0
```

4.  In 1–2 sentences: why is the z-test “more optimistic” than the
    t-test when $n$ is small?

    Because the z test assumes normal distrobution, allowing it to assume a bigger interval,
    as it therefore assumes that the standard deviation is known. This underestimated the true
    variability of the given distribution.

------------------------------------------------------------------------

## Exercise 5: One-sample chi-square test for a variance

Assume that `mtcars$mpg` is approximately normal. We test
$$
H_0:\sigma^2=\sigma_0^2\quad \text{with }\sigma_0^2 = 36
$$ (two-sided, $\alpha=0.05$).

Under $H_0$,
$$
Q=\frac{(n-1)S^2}{\sigma_0^2}\sim \chi^2_{n-1}.
$$

**Tasks**

1.  Compute $S^2$, $Q_{\text{obs}}$, and the two-sided p-value.

```{r}
S_squared <- sd(mtcars$mpg, na.rm=T)^2 
Q <- 31 * S_squared / 36
p_2sided <- 2 * min(
  pchisq(Q, df=31),      
  1 - pchisq(Q, df=31)  
)
```

```{r}
p_2sided
```

2.  Decide at $\alpha=0.05$. (Reject if in either tail.)

```{r}
p_2sided <= .05
```
The p-value is bigger then alpha $\rightarrow$ don't reject $H_0$


3.  State $H_0$ in words, and interpret the result in 1–2 sentences.

$H_0$: $\sigma_0^2 = 36$

Result: There is not enough evidence to reject $H_0$. This is the case, because the calculated
p-value is much bigger the given alpha, resulting in failure of disproving $H_0$

------------------------------------------------------------------------

# Part C — One-sample tests for a proportion (Bernoulli/binomial)

## Exercise 6: Proportion test (normal approximation vs exact)

We use the built-in table `Titanic` and focus on the overall survival
proportion.

Let $n$ = total passengers, $x$ = number of survivors, $\hat p=x/n$.
Test $H_0:p=0.38$ vs $H_1:p>0.38$ at $\alpha=0.05$.

```{r}
data(Titanic)
tab <- Titanic
n <- sum(tab)
x <- sum(tab[,,, "Yes"])
phat <- x / n
c(n = n, x = x, phat = phat)
```

**Tasks**

1.  Compute the normal-approximation z-statistic under $H_0$:
$$
    Z=\frac{\hat p-p_0}{\sqrt{p_0(1-p_0)/n}}.
    $$

```{r}
Z <- (phat - .38) / (sqrt(.38 * (1 - .38) / n))
```

2.  Compute the one-sided p-value $p=P(Z\ge z_{\text{obs}})$ and decide
    at $\alpha=0.05$.

```{r}
p_1sided <- 1 - pnorm(Z) 
p_1sided
```
Because the p-value is much bigger then alpha: 1 > 0.05, $H_1$ is not support and $H_0$
can definitely not be rejected.

3.  Compute the **exact** binomial p-value using `binom.test()`
    (one-sided, alternative = "greater"). Compare the two p-values and
    comment briefly.

```{r}
binom.test(x, n, .38, alternative = "greater")$p.value
```
identical.

4.  Check the rule of thumb for the normal approximation: are $np_0$ and
    $n(1-p_0)$ at least 10?

```{r}
n * .38 > 10
n * (1 - .38) > 10
```
$\rightarrow$ yes

------------------------------------------------------------------------

## Exercise 7: p-values for two-sided vs one-sided alternatives

Use your $t_{\text{obs}}$ from Exercise 2.

**Tasks**

1.  
```{r}
t_obs
```
<!-- -->

(a) two-sided $H_1:\mu\neq\mu_0$,

(b) right-tailed $H_1:\mu>\mu_0$,

(c) left-tailed $H_1:\mu<\mu_0$.

```{r}
#a 
2 * (1 - pt(abs(t_obs), df = 31))
#b
1 - pt(t_obs, df=31)
#c
pt(t_obs, df=31)
```

2.  Explain in 2–3 sentences why one-sided p-values are (typically)
    about half of two-sided p-values when $t_{\text{obs}}$ is in the
    direction of $H_1$.

The closer $t_{obs}$ is to $H_0$, the more equal both one-sided p-values become,
as the bigger one is getting smaller and the smaller one is getting bigger (trivial).
Because the 2-sided p-value is the p-value in both directions, it's getting closer
to $\text{left-sided-p} + \text{right-sided-p}$, which, if both are ~ equal, is
~ equal to $2 * \text{*-sided-p}$.

------------------------------------------------------------------------

# Part D — Type I/II error and power (simulation)

## Exercise 8: Power vs significance level (one-sample t-test)

We simulate a one-sample t-test for $H_0:\mu=\mu_0$ (two-sided). Assume
the data are normal with $\sigma=4$, $n=25$, and $\mu_0=20$.

We estimate: - $\alpha$ (Type I error) when the true mean is $\mu=20$, -
power when the true mean is $\mu=22$.

```{r}
sigma <- 4
n <- 25
mu_0 <- 20
```

**Tasks**

1.  Interpret the four numbers:
    -   Why is $\widehat\alpha$ close to the chosen $\alpha$?
    -   Why does power drop when you lower $\alpha$ from 0.05 to 0.01?

-

2.  (Conceptual) Explain why a one-sided test can have higher power than
    a two-sided test at the same $\alpha$. When is a one-sided test not
    justified?

-

------------------------------------------------------------------------

# Part E — Selected two-sample and chi-square methods (extension)

## Exercise 9: Two-sample means (pooled vs Welch) using `sleep`

The dataset `sleep` contains increases in sleep for two drugs.

```{r}
data(sleep)
head(sleep)
```

Let $X$ be group 1 and $Y$ group 2 (`extra` by `group`).

```{r}
x <- sleep$extra[sleep$group == 1]
y <- sleep$extra[sleep$group == 2]
c(n1 = length(x), n2 = length(y), mean1 = mean(x), mean2 = mean(y), sd1 = sd(x), sd2 = sd(y))
```

**Tasks**

1.  Perform Welch's two-sample test (default) and report the p-value.
2.  Perform the pooled two-sample t-test (`var.equal=TRUE`) and report
    the p-value.
3.  In 2–3 sentences: when would you prefer Welch over pooled?

```{r}
t.test(x, y, alternative = "two.sided", var.equal = FALSE)$p.value
```
```{r}
t.test(x, y, alternative = "two.sided", var.equal = TRUE)$p.value
```

I prefer the Welch test, as the pooled test needs very precise preconditions, like 
almost identical variance and n with reason (which is not given)

------------------------------------------------------------------------

## Exercise 10: Chi-square test of independence using `Titanic`

Test whether survival is independent of sex (ignoring other dimensions).

```{r}
tab2 <- apply(Titanic, c("Sex", "Survived"), sum)
tab2
```

**Tasks**

1.  State $H_0$ in words.
2.  Compute the chi-square test and interpret the p-value.
3.  Check whether expected counts are all at least 5.

$H_0$: Survival is independent of sex.


```{r}
ct <- chisq.test(tab2)
ct
```
p-value = $2.2e-16$

$\rightarrow$ The p-value is essentially zero, providing overwhelming evidence against the null hypothesis. We reject $H_0$ and conclude that survival is not independent of sex.

```{r}
ct$expected
```

All expected counts are above 5.
------------------------------------------------------------------------
