---
title: "Mock Exam — Statistics"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
  pdf_document: default
geometry: margin=1in
fontsize: 11pt
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
set.seed(123)
```

**Instructions**

-   Answer directly in this file and knit to HTML or PDF at the end.
-   Provide both **R output** and **short written interpretation** where
    requested.
-   Total: **130 points**.

------------------------------------------------------------------------

# Part A — R Basics and Probability Theory (40 points)

## A1. Data types and factors (7 points)

Create the following object and answer the questions.

```{r}
x <- c(1, 2, 2, 3, 3, 3)
g <- factor(c("A","B","A","B","B","A"))
df <- data.frame(x = x, g = g)
str(df)
```

1.  What does one **row** in `df` represent?

    *Write your answer here.*

2.  How many observations are in group `A` and in group `B`?\
    *Write your answer here.*

3.  Compute the **mean of `x`** separately for each group.

    ```{r}
    aggregate(df[1], FUN = mean, by = list(df$g))

    ```

## A2. Monte Carlo simulation (8 points)

Let $\Omega = \{1,2,\dots,100\}$. Draw $N=50{,}000$ values uniformly
from $\Omega$ (with replacement). Estimate:

$$
P(\text{multiple of 3} \mid \text{multiple of 3 or 5})
$$

1.  Compute the Monte Carlo estimate in R.

    ```{r}
    omega <- 1:100
    ssize <- 50000
    
    count <- 0
    success <- 0
    
    draws <- sample(omega, ssize, replace= TRUE)
    
    for(i in draws){
      
      if (i %% 3 == 0 || i %% 5 == 0){
        
        count <- count + 1
        
        if (i %% 3 == 0){
          
          success <- success + 1
          
        }
        
      }
        
    }
    
    success / count

    ```

2.  Compute the exact value by counting and compare.

    
    (P(i %% 3 == 0)) / (P(i %% 3 == 0) + P(i %% 5 == 0) - P(i %% 15 == 0))
    
    = 33 / (33 + 20 - 6) = 0.7010636
    
    monte carlo method result: 0.7018081
    
    delta_Results = abs(0.7018081 - 0.7010636) = 0.0007445
    
    => very precise awnser for such a context-loose question (P)
    

## A3. Sampling from a non-normal distribution (25 points)

The goal of this exercise is to illustrate the **Central Limit Theorem**
using simulation.

Let $X \sim \text{Exponential}(\lambda = 1)$.

1.  Draw $N = 10{,}000$ independent samples of size\
    $n = 5$, $n = 30$, and $n = 100$.

    ```{r}
    drawsize <- 10000
    
    X <- rexp(drawsize, 1)
    
    n5 <- sample(X , 5)
    
    n30 <- sample(X , 30)
    
    
    n100 <- sample(X , 100)
    
    ```

2.  For each $n$, compute the **sample mean**.

    ```{r}
    mean(n5)
    mean(n30)
    mean(n100)

    ```

3.  For each $n$, create a **histogram of the sample means** and overlay
    a **normal density** with the appropriate mean and variance.

    ```{r}
    hist(n5, probability = T)
    curve(dnorm(x, mean = mean(n5), sd = sd(n5)), add=T)
    
    hist(n30, probability = T)
    curve(dnorm(x, mean = mean(n30), sd = sd(n30)), add=T)
    
    hist(n100, probability = T)
    curve(dnorm(x, mean = mean(n100), sd = sd(n100)), add=T)

    ```

4.  For each sample size $n$, define the standardized mean

$$
Z_n = \frac{\bar X_n - \mu}{\sigma / \sqrt{n}},
$$

where $\mu$ and $\sigma$ are the true mean and standard deviation of the
exponential distribution.

a)  Compute $Z_n$ for $n = 5, 30, 100$.

    ```{r}
    
    compZ_n <- function(x_n, mu, sigma, n){
      return (x_n - mu) / (sigma / sqrt(n))
    }
    
    Z_5 <- compZ_n(mean(n5),1, 1, 5)
    Z_30 <- compZ_n(mean(n30),1, 1, 30)
    Z_100 <- compZ_n(mean(n100),1, 1, 100)
    

    ```

b)  Create histograms of $Z_n$ and compare them to the standard normal
    distribution.

    ```{r}
    hist(compZ_n(sample(X, n), mean(n), sd(n), n) + 2202, probability = T)
    curve(dnorm(x, mean = 1, sd = 1) * 2, add=T)
    ```

Answer the following:

c)  How does the distribution of $\bar X_n$ change as $n$ increases?

    It comes closer to the normal distribution

d)  Why does the CLT apply even though the original data are not
    normally distributed?

    Because:
    1. The sample size is big
    2. The chance to pull each value of the data is equally distributed

e)  What is the purpose of standardization in the CLT?

    It makes it possible to compare samples of different n's and lambda's, as
    they would normally have completely different means and distributions

------------------------------------------------------------------------

# Part B — Descriptive statistics (45 points)

We use `birthwt` from the MASS package.

```{r}
library(MASS)
library(dplyr)
library(ggplot2)
library(forcats)

data(birthwt)

bwt <- birthwt %>%
  mutate(
    smoke = factor(smoke, levels = c(0,1), labels = c("No","Yes")),
    ht    = factor(ht, levels = c(0,1), labels = c("No","Yes")),
    ui    = factor(ui, levels = c(0,1), labels = c("No","Yes")),
    low   = factor(low, levels = c(0,1), labels = c("No","Yes"))
  )
```

## B1. Variable overview (6 points)

1.  In your own words: what does one row of `bwt` represent?

    1 birth

2.  List **two categorical** and **two numeric** variables in `bwt`.

    cat: ht & smoke
    numeric: age & lwt

## B2. Histograms + ECDF (10 points)

For the numeric variable `bwt$bwt` (birth weight in grams):

1.  Create two histograms with different bin widths (e.g. 150 and 400).

    ```{r}
    hist(bwt$bwt, 5000 / 150)# -> gets rounded up to 200
    hist(bwt$bwt, 5000 / 400)# -> gets rounded up to 500
    
    
    ```

2.  Create an ECDF plot.

    ```{r}
    plot(ecdf(bwt$bwt))
    

    ```

3.  From the ECDF (or in R), estimate the proportion of babies with
    birth weight $\le 2500$ g. Write a sentence or two interpreting 1)
    how bin width affects the histogram and 2) what you see in the ECDF.

    ```{r}

    ```

    *Write your answer here.*

## B3. Group comparison: smokers vs non-smokers (9 points)

1.  Create a **boxplot** of `bwt` by `smoke`.

    ```{r}
    boxplot(bwt$bwt ~ bwt$smoke,
            xlab="Smoking (No / Yes)", ylab="Birth Weight", main = "How smoking during pregnancy relates to Birthweight")
    ```

2.  Report the median birth weight for each group.

    ```{r}
    mean_nonsmok <- mean(bwt$bwt[bwt$smoke == "No"], na.rm =T)
    mean_smok <- mean(bwt$bwt[bwt$smoke == "Yes"], na.rm = T)
    
    mean_nonsmok
    mean_smok
    ```

3.  In 3–4 sentences, summarize how smoking relates to birth weight
    using (i) the boxplot and (ii) the medians.

    ```{r}
     
    ```

    *Write your answer here.*

## B4. Contingency table + chi-square (10 points)

Consider the relationship between `smoke` and `low`.

1.  Create the contingency table.

    ```{r}
    library(dplyr)
    
    table(bwt$smoke, bwt$low)
    ```

2.  Compute **row-wise proportions** (within smoking status).

    ```{r}
    prop.table(table(bwt$smoke, bwt$low))
    ```

3.  Perform a chi-square test of independence and state the null
    hypothesis in words.

    ```{r}
    chisq.test(bwt$smoke, bwt$low)
    
    ```

    *Write your answer here.*

## B5. Scatterplot + correlation (10 points)

Study the relationship between maternal weight `lwt` and birth weight
`bwt`.

1.  Make a scatterplot (`bwt` vs `lwt`).

    ```{r}
    ggplot(bwt, aes(bwt, lwt)) +
      geom_point(size = 1) +
      labs(title = "birthweight vs maternal weight")
    ```

2.  Compute the correlation and interpret the sign and magnitude.

    ```{r}
    cor(bwt$bwt, bwt$lwt)
    ```
    Postive correlation (sign) -> the higher the bwt, the higher the lwt
    magnitude: the average lwt mapped to its bwt is lwt = bwt * 0.1857333

------------------------------------------------------------------------

# Part C — Inferential statistics (45 points)

We continue with the `birthwt` data (`bwt`) from Part B.

## C1. Linear regression inference + residual independence

Fit the simple linear regression model
$$
\texttt{bwt} = \beta_0 + \beta_1\,\texttt{lwt} + \varepsilon,
$$ where $\texttt{bwt}$ is birth weight (grams) and $\texttt{lwt}$ is
maternal weight (pounds).

### (a) Fit and summarize the model (5 points)

1.  Fit the model and print the summary output.

```{r}
    model <- lm(bwt$bwt ~ bwt$lwt)
    summary(model)
```

### (b) Point estimates and interpretation (5 points)

2.  Write down the estimated regression equation
    $\widehat{\texttt{bwt}} = b_0 + b_1\,\texttt{lwt}$.

        4.429 * lwt + 2369.624

3.  Interpret the slope $b_1$ **in context** (one sentence).

*Write your answer here.*

### (c) Confidence interval + hypothesis test for the slope (5 points)

We test whether there is evidence of a linear relationship between
maternal weight and birth weight.

4.  Compute a 95% confidence interval for $\beta_1$ (slope) using
    `confint()`.

```{r}
confint(model)

```

5.  Test $H_0:\beta_1=0$ vs $H_1:\beta_1\neq 0$ at $\alpha=0.05$. Report
    the observed $t$-statistic, the p-value, and your decision.

```{r}
sum_model <- summary(model)

t_stat <- sum_model$coefficients[2,3]
p_value <- sum_model$coefficients[2,4]

cat("t-Statistic:", t_stat, "\n")
cat("p-Wert:", p_value, "\n")
```

6.  Verify the CI–test duality: explain (in 1–2 sentences) why the
    decision in (5) agrees with whether $0$ lies inside/outside the 95%
    CI from (4).

*Write your answer here.*

### (d) Mean response CI vs prediction interval (5 points)

Let $x_0=130$ (maternal weight 130 lb).

7.  Use `predict()` to compute:
    -   a 95% confidence interval for the **mean** response
        $m(x_0)=E[\texttt{bwt}\mid \texttt{lwt}=x_0]$,
    -   a 95% **prediction** interval for a new observation
        $\texttt{bwt}_{new}$ at $\texttt{lwt}=x_0$.

```{r}
pred <- predict(model, newdata = data.frame(lwt = 130), interval = "confidence", level=.95)
colMeans(pred)
predict(model, newdata = data.frame(lwt = 130), interval = "prediction", level=.95) %>% colMeans
```
    CI: [2807.704 : 3081.470]


8.  Which interval is wider, and why? (1–3 sentences.)

*Write your answer here.*

### (e) Residual diagnostics (5 points)

9.  Extract the residuals $e_i$ and create:
    -   a plot of residuals versus observation index $i$,
    -   a Q-Q plot

```{r}

res <- residuals(model)

plot(res, main = "Residuals vs observation index", xlab = "observation index", ylab= "residuals", pch=19)
abline(h=0, col="red", lty=3)
```

## C2. Point estimation and confidence intervals (based on the lecture notes)

In this part you connect the regression work from C1 with the general
concepts from the lecture notes (estimators, bias/variance/MSE,
consistency, MLE, confidence intervals).

### (a) Population, sample, parameter, statistic (4 points)

1.  In the context of the variable `bwt$bwt` (birth weight in grams),
    describe:
    -   the **population** (conceptually),
    -   the **parameter** $\mu = \mathbb{E}[X]$,
    -   the **sample** $X_1,\dots,X_n$,
    -   one **statistic** $T(X_1,\dots,X_n)$.

    - all birth weights on earth
    - the true average / to be expected birth weight
    - the 189 birth samples in bwt
    - the expected birth weight / mean from the samples
    
### (b) Estimator vs estimate (2 points)

2.  Let $\bar X=\frac{1}{n}\sum_{i=1}^n X_i$. Explain the difference
    between the **estimator** $\bar X$ and the **estimate** $\bar x$.

    the estimator is the function / statistic, which returns an estimate upon recieving the
    necessary data, e.g. the lwt, race, etc.
    
    the estimate is the result of this

### (c) Bias, variance, MSE (4 points)

3.  Suppose an estimator $T$ of $\theta$ is unbiased and
    $\mathrm{Var}(T)=4$. Compute $\mathrm{MSE}(T)$.
    
    MSE = Bias² + Var
    MSE = 0² + 4 = 4

4.  Suppose another estimator $T'$ has $\mathrm{Bias}(T')=-1$ and
    $\mathrm{Var}(T')=1$. Compute $\mathrm{MSE}(T')$. Which estimator
    has smaller MSE?
    
    MSE = (-1)² + 1 = 2
    
    T' has the smaller MSE <!>

*Write your answer here.*

### (d) MLE in the uniform model (5 points)

5.  Assume $T_1,\dots,T_n \stackrel{iid}{\sim}\mathrm{Unif}(0,\theta)$
    and you observe

    beispieltext
    $$ t=(1.2,\ 0.7,\ 1.6,\ 0.9). $$

    Compute:

-   the MLE $\hat\theta_{\mathrm{MLE}}$,
-   the unbiased estimator $\tilde\theta=\frac{n+1}{n}\max_i t_i$.


<!>
```{r}
t= c(1.2,0.7, 1.6, 0.9)
n <- length(t)

MLE <- max(t)

unbest <- (n + 1) / n * MLE

MLE
unbest

```

*Write your answer here.*

### (e) Confidence interval for a mean + interpretation (5 points)

6.  Compute a 95% $t$-confidence interval for the mean birth weight
    $\mu$ using `t.test()`.

```{r}
t.test(bwt$bwt)$conf.int[1:2]

```

7.  Qualitative questions (no computation required):
    -   What happens to CI **length** if the confidence level increases
        from 95% to 99% (with the same data)?
    -   What happens to CI **length** if the sample size doubles (and
        the standard deviation stays roughly the same)?

*Write your answer here.*

# Submission checklist

-   [ ] The file knits without errors.
-   [ ] All plots have readable axis labels and titles.
-   [ ] Each requested interpretation is written in full sentences.
